---
title: "Google Certificate Capstone Study: An Exploration on Correlations between Stock Market Indexes and Cryptocurrencies"
author: "Andrés Gazzotti"
date: "2025-02-07"
output: html_document
---


### R Markdown
### Introduction
<div style="text-align: justify;">

##### In today´s "age of information", the democratization of access to finantial information has become a defining feature. The development of simple yet powerful mobile applications for executing trading operations has brought financial markets within reach of vast segments of the global population. While the question whether this new accessibity accessibility has increased market volatility lies outside the scope of this study 
##### Parallel to this shift, a new asset class emerged: *cryptocurrencies*. Bitcoin, introduced in early 2009 at a fraction of a cent in value, has evolved over sixteen years into what many consider a store of value. Cryptocurrencies now constitute a whole new market in its own right, with varying degrees of speculative activity across different tokens. This technological innovation has developed alongside mobile trading platforms, many of which now integrate both traditional markets (bonds, stocks, futures) and cryptocurrency options.
##### The following study is nothing more than an exploratory work on how the weekly closing prices of the traditional stock indexes (SP500, NASDAQ, Dow Jones) correlates with the same magnitude for bitcoin and ethereum in a span of 4 years, specifically between 01-01-2021 and 01-01-2025.


All the gathered data is publicly available at [Investing.com](https://www.investing.com/)

### R packages or libraries used in this code

#### library(lmtest)
#### library(reshape2)
#### library(dplyr)
#### library(tidyverse)
#### library(corrplot)
#### library(ggplot2)
#### library(GGally)
#### library(ggthemes)
#### library(vars)
#### library(patchwork)



```{r stock market and crypto, include=FALSE}
library(lmtest)
library(reshape2)
library(dplyr)
library(tidyverse)
library(corrplot)
library(ggplot2)
library(GGally)
library(ggthemes)
library(vars)
library(patchwork)
```
### Importing dataset
```{r}
merged_data <- read_csv("merged_data.csv")
```
#### Optional: Download the csv dataset here

```{r Optional downloading of the dataset} 
write.csv(merged_data, "merged_data.csv", row.names = FALSE)
```


### Creating the basic enviroment and dataset
```{r Variable Creation}
#Stock Market index first:
sp500 <- merged_data %>% 
  dplyr::select(Date, sp500)
nasdaq <- merged_data %>% 
  dplyr::select(Date, nasdaq)
dowjones <- merged_data %>% 
  dplyr::select(Date, dowjones)

#The two most prominent cryptocurrencies known as an storage of value on the market:
bitcoin <- merged_data %>% 
  dplyr::select(Date, bitcoin)
ethereum <- merged_data %>% 
  dplyr::select(Date, ethereum)
```
### Cleaning and exploration phase
### Let us check the structure of each data set
```{r Exploratory Phase}
#Stock market
str(sp500)
str(dowjones)
str(nasdaq)
#Crypto
str(bitcoin)
str(ethereum)
```

### Including Plots


```{r Exploratoy Plots}
#Making exploratory plots

#SP500
ggplot(sp500, aes(x = Date, y = sp500)) + geom_point(alpha = 0.5, size = 1.5) + 
  labs(title = "S&P 500 Closing Price 2021-2025", x = "Date by week", y = "Closing Price USD") + 
  theme_minimal()+
  theme_economist_white()
```
```{r DowJones}
#Dowjones

ggplot(dowjones, aes(x = Date, y = dowjones)) + 
  geom_point(alpha = 0.5, size = 1.5) + 
  labs(title = "DowJones Closing Price 2021-2025", x = "Date by week", y = "Closing Price USD") + 
  theme_minimal()+
  theme_economist_white()

```
```{r NASDAQ}
#Nasdaq
ggplot(nasdaq, aes(x = Date, y = nasdaq)) + 
  geom_point(alpha = 0.5, size = 1.5) + 
  labs(title = "Nasdaq Closing Price 2021-2025", x = "Date by week", y = "Closing Price USD") + 
  theme_minimal()+
  theme_economist_white()
```
```{r Bitcoin}
#Bitcoin

ggplot(bitcoin, aes(x = Date, y = bitcoin)) + 
  geom_point(alpha = 0.5, size = 1.5) + 
  labs(title = "Bitcoin Price 2021-2025", x = "Date by week", y = "Closing Price USD") + 
  theme_minimal()+
  theme_economist_white()
```
```{r Ethereum}
#Ethereum

ggplot(ethereum, aes(x = Date, y = ethereum)) + 
  geom_point(alpha = 0.5, size = 1.5) + 
  labs(title = "Ethereum Price 2021-2025", x = "Date by week", y = "Closing Price USD") + 
  theme_minimal()+
  theme_economist_white()
```

#### First insights: Nasdaq and SP500 share a similar tendency in their prices, while Dow Jones shows less consistency or similarity with the other two, this can be explain to the different industries each index is composed of and seems like SP500 and NASDAQ share an overlap.

### Now that we got our data sorted we can begin a basic statistical analysis first with the function summary: 

```{r Summary}
#SP500
summary(merged_data$sp500)
#DowJones
summary(merged_data$dowjones)
#Nasdaq
summary(merged_data$nasdaq)
#Bitcoin
summary(merged_data$bitcoin)
#Ethereum
summary(merged_data$ethereum)

```

##### Now we shall begin our correlation study running a test for each combination between the 3 stock market indexes an the two selected cryptocurrencies.

```{r Pearon Correlation Test: SP500 & Crypto}
Pn_sp500_Bit <- cor.test(merged_data$sp500, merged_data$bitcoin, method = "pearson")
print(Pn_sp500_Bit)
Pn_sp500_Eth <- cor.test(merged_data$sp500, merged_data$ethereum, method = "pearson")
print(Pn_sp500_Eth)

```

##### With a correlation pearson coefficient of 0.833617 for SP500 & Bitcoin and 0.6338 indicates that in both cases a moderate positive linear relationship exist between the prices of the stock index and the movements of the two analysed cryptocurrencies with a 95% confidence interval.

```{r Pearson Correlation Test: NASDAQ & Crypto}
Pn_Nas_Bit <- cor.test(merged_data$nasdaq, merged_data$bitcoin, method = "pearson")
print(Pn_Nas_Bit)
Pn_Nas_Eth <- cor.test(merged_data$nasdaq, merged_data$ethereum, method = "pearson")
print(Pn_Nas_Eth)
```
##### Analyzing the same test for the combinations with NASDAQ, we got an even stronger linear correlation than in the previous case, 0.90 for NASDAQ/Bitcoin and 0.719 for NASDAQ/Ethereum, given how NASDAQ is related to high tech capital this is expected. With a very low P-value of <2.2e^-16 in both cases being less than the significance label of 0.05 with a confidence interval of 95%.

```{r Pearson Correlation test: Dow Jones & Crypto}
Pn_Dow_Bit <- cor.test(merged_data$dowjones, merged_data$bitcoin, method = "pearson")
print(Pn_Dow_Bit)
Pn_Dow_Eth <- cor.test(merged_data$dowjones, merged_data$ethereum, method = "pearson")
print(Pn_Dow_Eth)
```
##### The Pearse coefficient -0.07 suggest a very weark negative correlation between Dow Jones prices and bitcoin prices; with a p-value being > than 0.05 and a confidence interval of 95%. In contrast with a correlation coefficient of 0.1675 a linear positive correlation exist although is a weak one, with a p-value > 0.05 it is statistically significant with a 95% confidence interval

##### Given these results we will use a heat map to better see them in accordance to its statistical relevancy 

```{r Pearson Correlation Heatmap}
heatmap_data_Pearson <- matrix(c(Pn_sp500_Bit$estimate, Pn_sp500_Eth$estimate,
                         Pn_Nas_Bit$estimate, Pn_Nas_Eth$estimate,
                         Pn_Dow_Bit$estimate, Pn_Dow_Eth$estimate), 
                       nrow = 3, ncol = 2, byrow = TRUE) %>%
  `rownames<-`(c("S&P 500", "NASDAQ", "Dow Jones")) %>%
  `colnames<-`(c("Bitcoin", "Ethereum")) %>% melt()

ggplot(data = heatmap_data_Pearson,aes(Var1, Var2, fill = value))+
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0.05, limit = c(-1, 1), space = "Lab", 
                       name="Correlation") +
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1)) +
  coord_fixed() +
  labs(title = "Pearson Correlation Heatmap Stocks & Crypto", x = "", y = "", 
       subtitle = "Weekly Prices from 2021-2025", caption = "Data collected from investing.com")
```

##### Given this linear correlation distribution we will study further this behavior with more graphs describing the correlative behavior between the series. Using the same gradient of colors as the last ilustration.

```{r}
ggplot(merged_data, aes(x = nasdaq, y = bitcoin)) +
  geom_point(alpha = 0.5, size = 1.5, color = "red") + 
  labs(title = "Nasdaq-Bitcoin Visual Correlation", x = "Nasdaq", y = "Bitcoin") + 
  theme_minimal()+
  theme_economist_white()

ggplot(merged_data, aes(x = nasdaq, y = ethereum)) +
  geom_point(alpha = 0.5, size = 1.5, color = "red") +
  labs(title = "Nasdaq-Ethereum Visual Correlation", x = "Nasdaq", y = "Ethereum") +
  theme_minimal()+ 
  theme_economist_white()

ggplot(merged_data, aes(x = sp500, y = bitcoin)) +
  geom_point(alpha = 0.5, size = 1.5, color = "red") +
  labs(title = "SP500-Bitcoin Visual Correlation", x = "SP500", y = "Bitcoin") +
  theme_minimal()+ 
  theme_economist_white()

ggplot(merged_data, aes(x = sp500, y = ethereum))+
  geom_point(alpha = 0.5, size = 1.5, color = "red") +
  labs(title = "SP500-Ethereum Visual Correlation", x = "SP500", y = "Ethereum") +
  theme_minimal()+ 
  theme_economist_white()
```

### Being those the most significant correlations (as per Pearson`s test). Now let us observe the lesser correlations accordingly with the first heatmap.

```{r}
ggplot(merged_data, aes(x = dowjones, y = bitcoin))+
  geom_point(alpha = 0.5, size = 1.5, color = "blue")+
  labs(title = "DowJones-Bitcoin Visual Correlation", x = "DowJones", y = "Bitcoin") +
  theme_minimal()+ 
  theme_economist_white()

ggplot(merged_data, aes(x = dowjones, y = ethereum))+
  geom_point(alpha = 0.5, size = 1.5, color = "orange")+
  labs(title = "DowJones-Ethereum Visual Correlation", x = "DowJones", y = "Ethereum") +
  theme_minimal()+ 
  theme_economist_white()
```

### Wtih these latter scatterplots it is clear that there is no strong correlation between the indicators.

### Given these results and to add more robustness to our analysis we will use another correlation test, one that is less sensitive to outlyers, the Kendall's Tau Correlation test.

### Kendall Tau Correlation Test:

#### Sp500 and crypto:

```{r}
Kt_sp500_Bit <- cor.test(merged_data$sp500, merged_data$bitcoin, method = "kendall")
print(Kt_sp500_Bit)
Kt_sp500_Eth <- cor.test(merged_data$sp500, merged_data$ethereum, method = "kendall")
print(Kt_sp500_Eth)
```

##### We get similar results as in the Pearson test but the coefficients are less pronounced 0.83 for Pearson and 0.61 for Kendall in sp500 & bitcoin and a 0.64 Pearson and 0.56 in Kendall's for the Sp500 ethereum correlation.

#### NASDAQ and crypto:
```{r}
Kt_Nas_Bit <- cor.test(merged_data$nasdaq, merged_data$bitcoin, method = "kendall")
print(Kt_Nas_Bit)
Kt_Nas_Eth <- cor.test(merged_data$nasdaq, merged_data$ethereum, method = "kendall")
print(Kt_Nas_Eth)
```
##### Same trend for the combinations of Nasdaq-Bitcoin and Nasdaq-Ethereum, with 0.72 and 0.60 coefficients respectively.


#### DowJones and crypto:

```{r}
Kt_Dow_Bit <- cor.test(merged_data$dowjones, merged_data$bitcoin, method = "kendall")
print(Kt_Dow_Bit)
Kt_Dow_Eth <- cor.test(merged_data$dowjones, merged_data$ethereum, method = "kendall")
print(Kt_Nas_Eth)
```

##### Surprisingly we get a more significant correlation coefficient than the past results with the Pearson's method specially concerning DowJones-Bitcoin, suggesting the possibility that outlyers were distorting the results, making the correlation less significant than it really was. This goes in contrast to the trend observed running the Kendall's Tau Coefficient methodology.

### Kendall Tau Correlation Heatmap:

##### With the same metodology as before, we will proceed to showcase the summarised results of this test on a heatmap.

```{r Kendall_Tau Heatmap}
heatmap_data_Kendall_Tau <- matrix(c(Kt_sp500_Bit$estimate, Kt_sp500_Eth$estimate,
                                 Kt_Nas_Bit$estimate, Kt_Nas_Eth$estimate,
                                 Kt_Dow_Bit$estimate, Kt_Dow_Eth$estimate), 
                               nrow = 3, ncol = 2, byrow = TRUE) %>%
  `rownames<-`(c("S&P 500", "NASDAQ", "Dow Jones")) %>%
  `colnames<-`(c("Bitcoin", "Ethereum")) %>% melt()

ggplot(data = heatmap_data_Kendall_Tau,aes(Var1, Var2, fill = value))+
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0.10, limit = c(-0.5, 1), space = "Lab", 
                       name="Correlation") +
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1)) +
  coord_fixed() +
  labs(title = "Kendall-Tau Correlation Heatmap Stocks & Crypto", x = "", y = "", 
       subtitle = "Weekly Prices from 2021-2025", caption = "Data collected from investing.com")
```

##### This general trend moderation of the correlation indicators under this method is something to be expected due to the qualities of the Kendall-Tau methodology being less sensitive to possible outlyers within the data. However, the results remain largely the same, expressing the same relationships between the variables.

##### As last correlation analysis we will dwell into the cross-correlation, this is nothing more than an examination of how similar are the time series, with normalized values.

### Cross-correlation:

```{r}
normalize <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}
normalized_data <- as.data.frame(lapply(merged_data[,-1], normalize))
normalized_data$Date <- merged_data$Date
```
#### Storing the time series as values:

```{r}
ts_sp500 <- ts(normalized_data$sp500)
ts_nasdaq <- ts(normalized_data$nasdaq)
ts_dowjones <- ts(normalized_data$dowjones)
ts_bitcoin <- ts(normalized_data$bitcoin)
ts_ethereum <- ts(normalized_data$ethereum)
```

##### After storing the values as a numerical vector of normalized values we take a look to the correlation between them as normalized time series

#### SP500 and crypto crossed correlation:
```{r}
ccf(ts_sp500, ts_bitcoin, main = "Cross-Correlation Plot SP500/BTC")
```

##### We see that the highest correlation coefficient is achieved at 0-lags at 0.83 as per Pearson test, alike our previous results. However it can be seen that this correlation gets weaker over time.

```{r}
ccf(ts_sp500, ts_ethereum, main = "Cross-Correlation Plot SP500/ETH")
```

##### On a similar note we see the same behavior of correlation here, getting the maximum, Pearson coefficient at 0-lags decreasing in statistical relevancy overtime.

#### NASDAQ and crypto crossed correlation.

```{r}
ccf(ts_nasdaq, ts_bitcoin, main = "Cross-Correlation Plot Nasdaq/BTC")
```

##### Remembering the highest correlation recorded in this study the Pearson coefficient of 0.90 is achieved at 0-lags decreasing over time at less than 0.60 by the lag 20.

```{r}
ccf(ts_nasdaq, ts_ethereum, main = "Cross Correlation Nasdaq/ETH")
```

##### Same behavior regarding lags in the time series, with slightly more sudden decrease in statistical relevancy after the apex of the value at 0-lags.

#### Dow Jones and crypto crossed correlation:

```{r}
ccf(ts_dowjones, ts_bitcoin, main = "Cross Correlation DowJones/BTC")
```

##### Recalling the lowest linear correlation between Dow Jones weekly closing prices and the same for Bitcoin, recorded a 0.08 Pearson coefficient. However in the graph we can see that at times it presents a negative correlation between lag at {-20, -13} and then again at {-7, 3} lags, the correlation is only statistically relevant at {4, ∞+) on wards.This inconsistency shows an overall weak correlation between the two and sensitive to outlyers like we saw with using less sensitive tests.

```{r}
ccf(ts_dowjones, ts_ethereum, main = "Cross Correlation DowJones/ETH")
```

##### Differently to other combinations we see that both combination on Dow Jones are not stable, through the series we see and increase and subsequent decrease in statistical relevancy with two spikes from {-19, -4} and then again from {-1, 17}. Overall the correlation is weak but not as weak as the Dow Jones/Bitcoin combination.

#### Summarizing plot using corr_plot package.

```{r}

library(corrplot)
#Removing date:
norm_for_corrplot <- normalized_data %>% 
  dplyr::select(-Date)

corrplot(cor(norm_for_corrplot),
         method = "number",
         type = "upper")
```

##### In this last plot we can see, for reference the linear correlation between the stock indexes as well.

### Correlation does _not_ means causation

##### As one of the most well known phrases of econometric it is worth remembering that correlation does not means that one variable or time series cause the other, it could be a link to thing in that direction but by itself does not prove by itself the causation relationship. Various external factors and complex dynamics can influence these relationships, and further investigation is required to establish any causal connections.

### Granger Causality Test

```{r}
# Example of loading data
ts_data <- cbind(ts_bitcoin, ts_nasdaq, ts_dowjones, ts_ethereum, ts_sp500)
# Select the optimal number of lags using AIC and BIC criteria
lag_selection <- VARselect(ts_data, lag.max = 20, type = "const")

# Saving the optimal lag
print(lag_selection)
optimal_lags <- lag_selection$selection["AIC(n)"]
optimal_lags <- as.numeric(optimal_lags)
```

#### NASDAQ and Bitcoin

```{r}
granger_nas_bit <- grangertest(ts_bitcoin ~ ts_nasdaq, order = optimal_lags) %>% 
  print()
granger_bit_nas <- grangertest(ts_nasdaq ~ ts_bitcoin, order = optimal_lags) %>% 
  print()
```

##### With a p-value below 0.05 it suggests that prices of NASDAQ can have a causation effect on bitcoin, however the same thing cannot be said the other way around as bitcoin prices don't seem to have a causation effect on NASDAQ prices p-value > 0.05.

#### NASDAQ and Ethereum.

```{r}
granger_nas_eth <- grangertest(ts_ethereum ~ ts_nasdaq, order = optimal_lags) %>% 
  print()
granger_eth_nas <- grangertest(ts_nasdaq, ts_ethereum, order = optimal_lags) %>% 
  print()
```

##### With a p-value < 0.01 there are reasons to believe that there is some causation in both directions.

#### SP500 and Bitcoin
```{r}
granger_sp500_bit <- grangertest(ts_bitcoin ~ ts_sp500, order = optimal_lags) %>% 
  print()
granger_bit_sp500 <- grangertest(ts_sp500, ts_bitcoin, order = optimal_lags) %>% 
  print()
```

##### In both directions with a P-value > 0.01, the null hypothesis is approved that there is not significant evidence that suggest causation.

#### SP500 and Ethereum

```{r}
granger_sp500_eth <- grangertest(ts_ethereum ~ ts_sp500, order = optimal_lags) %>% 
  print()
granger_eth_sp500 <- grangertest(ts_sp500 ~ ts_ethereum, order = optimal_lags) %>% 
  print()
```

##### Just like the last combination the results suggest that there are not evidence to suggest a causation relation between the variables in both directions. With a P-value of 0.08302 causation ethereum-sp500 and a P-value of 0.4276 for Sp500-ethereum.

#### DowJones and Bitcoin

```{r}
granger_dowj_bit <- grangertest(ts_bitcoin ~ ts_dowjones, order = optimal_lags) %>% 
  print()
```
##### Very weak link for causation of the Dow Jones prices to cause bitcoin prices, in line with the weak correlation between the two with a P-value < 0.05 but at a weak presentation 0.04127.

```{r}
granger_bit_dowj <- grangertest(ts_dowjones ~ ts_bitcoin, order = optimal_lags) %>% 
  print()
```

##### With a P-value at 0.5402 implies P-value > 0.05, with evidences for not refusing the null hypothesis implying that past values of bitcoin does not provide significant information on predicting Dow Jones prices.

#### DowJones and Ethereum:

```{r}
granger_dowj_eth <- grangertest(ts_ethereum ~ ts_dowjones, order = optimal_lags) %>% 
  print()
```

##### Same result regarding past values of Dow Jones on providing relevant information for prices of Ethereum, contrary to the little correlation that was shown, with a 0.18 P-value, the P-value > 0.05, there are evidence to disrpove a relation of causality.

```{r}
granger_eth_dowj <- grangertest(ts_dowjones ~ ts_ethereum, order = optimal_lags) %>% 
  print()
```

##### With a P-value of 0.9261 the result of the test is well above P-value > 0.05. It can be disproven any relevant relationship of causality.

### Conclusions

##### The stock market and the most valuable cryptocurrencies have shown interesting correlations between them, for NASDAQ and SP500 the results were more significant results than Dow Jones, fact that can be linked with the composition of the industries within said index. NASDAQ, very related with the so called "Big Tech" makes it logically more related to newer systems like crypto while more traditional stocks like those in Dow Jones are less  related. SP500 is a interesting case, the stock index is used many times as a "general picture" for the american economy, the relation between the index and crypto was always present suggesting the growing importance of technological stocks within this index. We can see this growing importance of crypto and connection between this financial markets with the creation of ETF (Exchange-Traded Fund) for both Bitcoin and Ethereum last year listed in NASDAQ and other stock index.

##### It is worth mentioning that further studies need to be made, the objective of this study was an exploratory focus for the possible links between both traditional markets and newer less regulated markets, it does not suggest a direct link or the nature of the relation between them. Further methods need to be applied considering other variables into a wider model, considering inflation expectations, interest rate and other variables. Also it is recommended the decomposition of variables into a more complex auto regressive model such as VAR, SVAR among others. 


### References

##### Correlation coefficient and correlation test in R. (n.d.). Stats and R. https://statsandr.com/blog/correlation-coefficient-and-correlation-test-in-r/
##### Investing.com - stock market quotes & financial news. (n.d.). Investing.com. https://investing.com/
##### Kendall’s Tau - StatsTest.com. (2020, April 7). StatsTest.com. https://www.statstest.com/kendalls-tau/
##### Statistics Solutions. (2025, January 17). Understanding Pearson’s Correlation Coefficient: A Guide to Statistical Association. https://www.statisticssolutions.com/free-resources/directory-of-statistical-analyses/pearsons-correlation-coefficient/


</div>

